Estimating dark matter halo concentration using the integrated mass profile

### 2016-04-17 paper/paper.tex
lots of new text in the paper

\begin{abstract}
We present a new algorithm to estimate the concentration of N-body
dark matter halos using the integrated mass profile.
The method uses the full particle information without any binning,
making it reliable in cases when low numerical resolution becomes a
limitation for other methods.   
We test the performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning.
Tests on the mock halos show that the accuracy of the new method to
recover known input concentrations varies with halo resolution,
outperforming the other two methods. 
We also measure the mass-concentration relationship on N-body
data. 
We find that in the probed mass range ($10^{12}\hMsun <
10^{14}\hMsun$) the three methods give consistent results within the
statistical uncertainties.   
We only find a small deviation at low masses, $M<10^{13}\hMsun$, where
the new method yields lower median concentration values by $20\%-30\%$
compared to the velocity and density methods. 
From these results we believe that the new method is a promising tool
to probe the internal structure of dark matter halos.
\end{abstract}

@@

In this letter we present a third alternative based on fitting the
integrated mass profile.
This approach has two advantages with respect to the above mentioned
methods.  
It does not involve any data binning and does not throw away data
points. 
This translates into a robust estimate even at low resolution/particle
numbers.   
Furthermore, since the method does not require any binning, there is
no need to tune numerical parameters.
This method provides a new independent method to estimate the
concentration parameter.   
Here we provide a detailed description of the method and a
comparison against the two traditional techniques by measuring its
performance both on mock and N-body halos.

@@

In this letter we use $\Delta_h=740$, a number roughly corresponding to
200 times the critical density at redshift z=0.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.45\textwidth]{error_conc_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number $N$ in
  the set of mock halos. Different symbols represent different
  methods. The new method provides the most accurate estimate at fixed
  particle number.  
    \label{fig:downsampling}}
\end{figure}

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.45\textwidth]{concentration_bolshoi.pdf}
  \includegraphics[width=0.45\textwidth]{concentration_bolshoi_corrected.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data. 
  All halos are used regardless of their relaxation state.
  The lines correspond to the median concentration values in each mass bin.
  The shaded region presents 10 and 90 per cent spread.
  For clarity we only show the spread for the concentrations estimated
  using the new method. The other two methods have a similar spread. The
  dotted line corresponds to fits reported by \citep{Prada2012}.
  \label{fig:concentration}}
\end{figure*}

@@

The line on the same panel shows a fit to the data with the following
functional form 
\begin{equation}
f_{\rm off} = \frac{A}{(1+\log_{10}N_s)^{B}}.
\end{equation} 
for the dataset shown in the panel we have $A=3385\pm 2065$, $B=7.99\pm0.48
$ for the velocity method and $A=-110\pm 253$, $B=7.11\pm1.80$ for the
new method, where the uncertainties are estimated from the covariance 
matrix resulting from the non-linear optimization procedure. The lines
on Figure \ref{fig:downsampling} are plotted using these best values.

Figure \ref{fig:downsampling} summarizes different uncertainty
results. 
Large symbols show the standard deviation on $f_{\rm off}$
from the downsampling experiment.
Small dots show the uncertainty on each individual concentration value
as estimated by the MCMC algorithm.
The line shows the $\propto N^{-1}$  locus.
The first thing to notice is that statistical uncertainties are at
least one order of magnitude than the uncertainty associated to
downsampling experiment.

@@

Figure \ref{fig:concentration} summarizes the previous results in the
mass-concentration relationship. 
We find that the median concentration
can be fit by the following relationship
%\begin{equation} c_{200} = 5.62 \left(\frac{M_{200}}{10^{12}\hMsun}\right)^{-0.061}, 
%\end{equation}
%
which has a similar power value compared to the $-0.075$ reported by
\citep{Prada2012}, while the normalization is $\sim 20\%$ lower
compared to the $7.28$ value reported by the same authors.

@@

In this letter we presented a new method to estimate the concentration
of dark matter halos in N-body simulations.  
We tested the method on mock halo data to study the impact of total
number of particles and input concentration on the retrieved values.  
We compared these results against two other methods commonly used in
the literature to estimate concentrations.  
For all methods, the accuracy in retrieving the input concentration
increases with the number of particles as summarized in Figure
\ref{fig:error}.  The new method systematically outperforms the other two.

We also applied the method to halos extracted from a cosmological
N-body simulation to estimate the impact on the mass concentration
relationship. Although the three  methods are in broad agreement there
are some noticeable differences in the mass concentration relationship
at lower masses, $<10^{13}\hMsun$, at the $20-30\%$ level.

@@

\vspace{0.5cm}

>>>>

\begin{abstract}
We present a new algorithm to estimate the concentration of N-body
dark matter halos using the integrated mass profile.
The method uses the full particle information without any binning,
making it reliable in cases when low numerical resolution becomes a
limitation for other methods.   
We test the performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning.
Tests on the mock halos show that the accuracy of the new method to
recover known input concentrations varies with halo resolution,
outperforming the other two methods. 
Convergence tests by downsampling halos with high number of particles in
the N-body simulation show that the new method is robust with respect
to changes in halo particle number.
However the velocity method overpredicts concentration values by a
factor of $25\%$ values for low particle numbers.
We also measure the mass-concentration relationship on N-body
data. 
We find that in the probed mass range ($10^{10}\hMsun < M_{h} <
10^{14}\hMsun$) the three methods give consistent results within the
statistical uncertainties.   
We find a small deviation at low masses, $M<10^{12}\hMsun$, where
the new method yields lower median concentration values by $25\%$
compared to the velocity methods. 
We find that accounting for the bias
in the velocity method brings the results into agreement.
From these results we believe that the new method is a promising tool
to probe the internal structure of dark matter halos.
\end{abstract}

@@

In this paper we present a third alternative based on fitting the
integrated mass profile.
This approach has two advantages with respect to the above mentioned
methods.  
It does not involve any data binning and does not throw away data
points. 
This translates into a robust estimate even at low resolution/particle
numbers.   
Furthermore, since the method does not require any binning, there is
no need to tune numerical parameters.
This method provides a new independent method to estimate the
concentration parameter.   
Here we provide a detailed description of the method and a
comparison against the two traditional techniques by measuring its
performance both on mock and N-body halos.

@@

In this paper we use $\Delta_h=740$, a number roughly corresponding to
200 times the critical density at redshift z=0.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.45\textwidth]{error_conc_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number $N$ in
  the set of mock halos. Different symbols represent different
  methods. The new method provides the most accurate estimate at fixed
  particle number.  
    \label{fig:downsampling_err}}
\end{figure}

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.45\textwidth]{concentration_bolshoi.pdf}
  \includegraphics[width=0.45\textwidth]{concentration_bolshoi_corrected.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data. 
  All halos are used regardless of their relaxation state.
  The lines correspond to the median concentration values in each mass bin.
  The shaded region presents 10 and 90 per cent spread.
  For clarity we only show the spread for the concentrations estimated
  using the new method. The other two methods have a similar spread. The
  dotted line corresponds to fits reported by \citep{Prada2012}.
  \label{fig:concentration}. The left panel shows the raw results
  comming from each algorithm. The right panel introduces a
  correction on the velocity and integrated mass results. 
  The corrections dependens on the number of particles following the
  results of the downsampling experiments presented in Figure
  \ref{fig:downsampling}.} 
\end{figure*}

@@

The line on the same panel shows a fit to the data with the following
functional form 
\begin{equation}
f_{\rm off} = \frac{A}{(1+\log_{10}N_s)^{B}}.
\label{eq:f_off}
\end{equation} 
for the dataset shown in the panel we have $A=3385\pm 2065$, $B=7.99\pm0.48
$ for the velocity method and $A=-110\pm 253$, $B=7.11\pm1.80$ for the
new method, where the uncertainties are estimated from the covariance 
matrix resulting from the non-linear optimization procedure. The lines
on Figure \ref{fig:downsampling} are plotted using these best values.

Figure \ref{fig:downsampling} summarizes different uncertainty
results. 
Large symbols show the standard deviation on $f_{\rm off}$
from the downsampling experiment.
Small dots show the uncertainty on each individual concentration value
as estimated by the MCMC algorithm.
The line shows the $\propto N^{-1}$  locus.
This figure demonstrates that the statistical uncertainties are at
least one order of magnitude than the uncertainty associated to
downsampling experiment.

@@

Figure \ref{fig:concentration} shows the  mass-concentration
relationship for the density, velocity and integrated mass method.  
The left panel shows the results as they are produced by each of the
algorithms, the thin dashed line marks the results by \citep{Prada2012}.

We notice first that the results from the density method have a
systematic $15\%$offset from the velocity methods.
This offset was already presented by \citep{Prada2012} for some
concentrations and halo masses, although a more organized study in
terms of the mass-concentration relationship has not been presented in
the literature.

We then notice that our methods follows closely the results of the
velocity method at higher masses, $M_h>10^{12}\hMsun$ or equivalently
for $>5\times10^3$ particles. 
For masses below that threshold our method saturates at a
concentration value of $\sim 7$ while the velocity method continues to
give higher concentration values.

We hypothesize that the increase in the results for the velocity
method below $5\times 10^{4}$ particles come from the systematic
bias described in the previous subsection.  
To test the general consistency of this hypothesis wwe correct the
concentration values in the velocity and integrated mass methods by
a factor of $1/(1+f_{\rm off})$, using the definition in Equation
(\ref{eq:f_off}) and the parameters obtained from the data presented
in Figure (\ref{fig:downsampling}).   
The correction brings into perfect agreement the results between the
velocity and the integrated mass method. 
This suggests that the concentration at lower masses might very well
reach a plateu in the halo mass range $10^{10}< M_{h}/\hMsun < 10^{12}$ .

@@

In this paper we presented a new method to estimate the concentration
of dark matter halos in N-body simulations.  
We tested the method on mock halo data to study the impact of total
number of particles and input concentration on the retrieved values.  
We compared these results against two other methods commonly used in
the literature to estimate concentrations.  
For all methods, the accuracy in retrieving the input concentration
increases with the number of particles as summarized in Figure
\ref{fig:error}.  The new method systematically outperforms the other two.

We also applied the method to halos extracted from a cosmological
N-body simulation.
Using this data we showed that the new method is robust with respect to a
decreasing number of particles in the halos. 
On the other hand we showed that the velocity method has a
systematic bias that overesimates the concentration for halos with a
low particle number up to a factor of $25\%$.

We also used the N-body simulation to estimate the impact on the mass
concentration relationship. 
Although the three methods are in broad agreement within their
uncertainties there are some noticeable differences.
The first difference is that the density method produces
systematically higher concentrations by a factor of $15\%$ compared to
the velocity mehod.
The second difference is that the velocity and integrated mass methods
start to differ for masses below $10^{12}\hMsun$. 
We found that a correction of the results by the bias factor found
using the same N-body data can bring these two methods into
agreement.

@@

\vspace{0.1cm}
